Array

### 2013-12-23 DOCUMENTATION/AdvancedTypes.tex
Intermediate commit (reorganizing repo).

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\ensuremath{\downarrow}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\lbl}{\mathcal{L}()}
\DeclareMathOperator{\variable}{\mathcal{V}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[0]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\developed}[0]{\ensuremath{\textsf{developed}}}
\newcommand{\stateless}[0]{\ensuremath{\textsf{stateless}}}
\newcommand{\importedmodules}[0]{\ensuremath{\textsf{i-mod}}}
\newcommand{\trimportedmodules}[0]{\ensuremath{\textsf{i-mod$\mathsf{\ast}$}}}
\newcommand{\importedinstances}[0]{\ensuremath{\textsf{i-inst}}}
\newcommand{\trimportedinstances}[0]{\ensuremath{\textsf{i-inst$\mathsf{\ast}$}}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Theta}}
\newcommand{\stateref}[0]{\ensuremath{\sf\Phi}}
\newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\init}[0]{\ensuremath{\mathsf{I}}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}

@@

\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
  arr(2):=5;
  i = arr(2);
\end{minted}
\end{footnotesize}

\begin{footnotesize}
\begin{minted}[frame=single,bgcolor=bg,linenos=true]{llvm}
  store i32 5, i32* getelementptr  ([100 x i32]* @arr, i32 0, i64 2)
  %1 = load i32* getelementptr  ([100 x i32]* @arr, i32 0, i64 2)
  store i32 %1, i32* @i, align 4
\end{minted}
\end{footnotesize}

%TODO update the text showing that to get the address is needed to get a 
%sequence of steps.

@@

\iffalse 
\paragraph{An atomic expression}

\begin{align*}
\lefteqn{\trad{n}{\B{Name}} \isdef \LET t = \trad{n.type}{\B{Type}} \IN} \\
& \IF \B{n} \mbox{ is a constant } \THEN \trad{n.value}{\B{Expr}} \\
& \ELSIF \B{n} \mbox{ is a local variable } \THEN
\llvm{""} \sep \local(\B{n}) \sep t \\
& \ELSE (\B{n} \mbox{ is a primitive state variable }) \LET p = \variable \AND v = \variable \IN \\
& \quad \llvm{"\PH{p} = getelementptr  \PH{\stateref(\B{n.root})} \%self\$, i32 0, i32 \PH{\idx(\B{n})} \nl} \\
& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t\\
& \ELSE (\B{n} \mbox{ is an array}) \LET p = \variable \AND v = \variable \IN \\
& \quad \llvm{"\PH{p} = getelementptr \PH{\state(\B{n.root})} \%self\$,
\PH{\stateindex(n)} \nl} \\
& \quad \llvm{\PH{v} = load \PH{t}* \PH{p}\nl"} \sep v \sep t
\\
\end{align*}
\fi

\paragraph{Becomes equal instructions} must evaluate the target and the source
of the assignment, and then copy the result of the latter in the former, using
the LLVM instruction \llvm{store}. The evaluation of the target is defined with
$\trad{}{lvalue}$, which yields the corresponding LLVM source code $l$ and the
variable $p$ containing the assigned location. The evaluation of the source
yields the corresponding code $r$, the variable $v$ holding the result, and its
type $t$. \VGM{Todo: update only the function LValue}
\begin{align*}
\begin{split}
  \trad{a}{\B{Beq}} \isdef
  & \LET l \sep p \sep t'= \trad{a.lhs}{lvalue} \AND r \sep v \sep t = \trad{a.rhs}{\B{Expr}} \IN \\
  & \llvm{"\PH{l}} \\
  & \llvm{\PH{r}} \\
  & \llvm{store \PH{t} \PH{v}, \PH{t'} \PH{p} \nl"}
%   & l \conc r \conc
%   \llvm{"store "} \conc t \conc \texttt{" "} \conc v \conc \llvm{", "}
%   \conc t \conc \llvm{"* "} \conc p \conc \nl
\end{split}
\end{align*}

TODO: This function updates the Name function ?
% Rule to get a element in array, it can be used in become equal or Expressions
\begin{align*}
\lefteqn{\trad{n}{lvalue} \isdef} \\
& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}}"} \IN \\
& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
& \quad \IF \mbox{\B{n} is an array} \THEN \LET v = \variable \IN \trad{n}{ArrayItem}  \sep v \sep t \\
& \quad \ELSE \LET v = \variable \IN \\
& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
\end{align*}

% We cannot use the tal function, because it is not able to solve any 
% expressions.

The rule ArrayItem is applied in expressions as ``$a(n1,...,n_{n})$'', where $a$ is n.base 
and $n1,...,n_{n}$ is a sequence of indices to select an item from array.

@@

The assignment target may be either local to the operation or a state
variable. In the former case, it can be referenced by the corresponding LLVM
identifier. In the latter case, it is represented as an element of the state
structure $\Sigma$, and its address must be calculated and assigned to a new temporary
variable.

\paragraph{Definition of function:} $\stateindex(\B{index,type}) \isdef$ to get a 
sequence of index to select the element. If $\B{n} $ is integer vector ``arr'' 
with 100 elements selecting the element 1 then the result is this text ``, i32 
0, i64 1''. It means to select the element in dimension zero and position one of 
array, where the index types are respectively ``i32'' and ``i64''. More detailed 
examples are presented in the reference manual of LLVM-IR. %TODO: put her a 
% reference of manual

@@

\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
 array1 := array2;
 structure1 := structure2;
\end{minted}

@@

%TODO update this example
\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
Year : struct(
	Year_number : INT,
	Leap_year : BOOL,
	Number_of_days : (1..12) * (28..31),
	Weather : struct( Average_Temperature : (1..12) *  INT,
		Average_Rain : (1..12) * INT) )
\end{minted}

The following code declares a record example:

\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
Year := rec(
	Year_number : 2013,
	Leap_year : TRUE,
	Number_of_days : (12,28),
	Weather : rec(  Average_Temperature : (5,8),
			Average_Rain : (5,10)))
\end{minted}

This code stores a value in the record and get the value of element.

@@

\subsection{A  simple example and its LLVM-IR}
\footnotesize{
\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
MACHINE
   Array
VARIABLES arr

@@

\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
IMPLEMENTATION 
   Array_i
REFINES
   Array
CONCRETE_VARIABLES
   arr
INVARIANT 
    arr :  ((0..99) --> 0..1000)
INITIALISATION
   arr(1) := 100
OPERATIONS
   set ( ix , tt ) =
   IF
       ix >= 0 & ix <=99 & tt >= 0  & tt <= 1000
   THEN

@@

\begin{minted}[frame=single,bgcolor=bg,linenos=true]{llvm}

@@

}

@@

\begin{footnotesize}
\begin{minted}[frame=single,bgcolor=bg,linenos=true]{pascal}
MACHINE
   Bubble
VARIABLES
   vec,sort
INVARIANT
   vec : 0..99 --> 0..99 & sort : 0..1 &
   ( sort = 1 =>  !ii.(ii : 0..98 => vec(ii)<=vec(ii+1)))
INITIALISATION
   vec :: 0..99 --> 0..99 || sort := 0
OPERATIONS
   op_sort =
   ANY sorted_vector WHERE
    sorted_vector : 0..99 --> 0..99 &
    !ii.(ii : 0..98 => sorted_vector(ii)<=sorted_vector(ii+1))
   THEN
    vec := sorted_vector|| sort := 1
   END

@@

\end{minted}
\end{footnotesize}

>>>>

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\ensuremath{\downarrow}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\lbl}{\mathcal{L}()}
\DeclareMathOperator{\variable}{\mathcal{V}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[0]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\developed}[0]{\ensuremath{\textsf{developed}}}
\newcommand{\stateless}[0]{\ensuremath{\textsf{stateless}}}
\newcommand{\importedmodules}[0]{\ensuremath{\textsf{i-mod}}}
\newcommand{\trimportedmodules}[0]{\ensuremath{\textsf{i-mod$\mathsf{\ast}$}}}
\newcommand{\importedinstances}[0]{\ensuremath{\textsf{i-inst}}}
\newcommand{\trimportedinstances}[0]{\ensuremath{\textsf{i-inst$\mathsf{\ast}$}}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Theta}}
\newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\init}[0]{\ensuremath{\mathsf{I}}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}

@@

\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
  arr(2):=5;
  i = arr(2);
\end{minted}
\end{footnotesize}

\begin{footnotesize}
\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{llvm}
  store i32 5, i32* getelementptr  ([100 x i32]* @arr, i32 0, i64 2)
  %1 = load i32* getelementptr  ([100 x i32]* @arr, i32 0, i64 2)
  store i32 %1, i32* @i, align 4
\end{minted}
\end{footnotesize}

TODO: show a simple example getting or setting an element of array 
multidimensinal using the rule $ArrayItem$. For example, show the translation of 
this code: $ arr(2,3) :=4$.

@@

\paragraph{Becomes equal instructions} must evaluate the target and the source
of the assignment, and then copy the result of the latter in the former, using
the LLVM instruction \llvm{store}. The evaluation of the target is defined with
$\trad{}{lvalue}$, which yields the corresponding LLVM source code $l$ and the
variable $p$ containing the assigned location. The evaluation of the source
yields the corresponding code $r$, the variable $v$ holding the result, and its
type $t$.

\begin{align*}
\begin{split}
  \trad{a}{\B{Beq}} \isdef
  & \LET l \sep p \sep t'= \trad{a.lhs}{lvalue} \AND r \sep v \sep t = \trad{a.rhs}{\B{Expr}} \IN \\
  & \llvm{"\PH{l}} \\
  & \llvm{\PH{r}} \\
  & \llvm{store \PH{t} \PH{v}, \PH{t'} \PH{p} \nl"}
\end{split}
\end{align*}

% Rule to get a element in array, it can be used in become equal or Expressions
\begin{align*}
\lefteqn{\trad{n}{lvalue} \isdef} \\
& \LET t = \llvm{"\PH{\trad{\B{n.type}}{Type}}"} \IN \\
& \quad \IF \mbox{\B{n} is a local variable} \THEN \llvm{""} \sep \local(\B{n}) \sep t \\
& \quad \IF \mbox{\B{n} is an array} \THEN  \trad{n}{ArrayItem} \\
& \quad \ELSE \LET v = \variable \IN \\
& \quad \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{n.root})}* \%self\$,} \\
& \quad \quad \quad \llvm{i32 0, i32 \PH{\idx(\B{n})} \nl"} \sep v \sep t
\end{align*}

\paragraph{Array item} selects an item in expressions as ``$a(n1,...,n_{n})$'', 
where $a$ is $n.base$ and $n1,...,n_{n}$ is a sequence of indices that 
calculates the address of element.

@@

@@

\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
 array1 := array2;
 structure1 := structure2;
\end{minted}

@@

\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}

MACHINE Rec
SETS
 ID={aa,bb}
VARIABLES xx, bank, total
INVARIANT
 xx : struct(name : ID, balance : NAT) & bank : struct(name : ID) &
 total:struct(money:NAT)
INITIALISATION 
 bank := rec(name:bb) ||
 xx:= rec(name:aa, balance:10) ||
 total := rec(money:20)
OPERATIONS

  res<--positive = PRE xx'balance > 0 THEN res:=TRUE END;

  withrdaw(amt) = PRE amt:NAT1 & (xx'balance >= amt) THEN
     xx := rec(name:xx'name, balance:(xx'balance - amt))
  END;

  unsafe_dec = BEGIN
       xx'balance := xx'balance - 1
  END
END

\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
IMPLEMENTATION
    Rec_i

REFINES
    Rec

CONCRETE_VARIABLES
   xx ,
   bank ,
   total

INITIALISATION

   bank := rec(name:bb) ;
   xx := rec(name:aa, balance:10) ;
   total :=rec(money:20)
OPERATIONS

  res<--positive = 
  IF xx'balance > 0 THEN
       res:= TRUE 
  ELSE res := FALSE
  END;

  withrdaw(amt) = IF   amt >=0 & amt <= MAXINT & (xx'balance >= amt) THEN 
     xx := rec(name:xx'name, balance:(xx'balance - amt))
  END;

  unsafe_dec = BEGIN
       xx'balance := xx'balance -1 
  END
END
\end{minted}

@@

\subsection{A  simple example and its LLVM-IR}

\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}

MACHINE
   Array
VARIABLES arr

@@

\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
IMPLEMENTATION 
   Array_i
REFINES
   Array
CONCRETE_VARIABLES
   arr
INVARIANT 
    arr :  ((0..99) --> 0..1000)
INITIALISATION
   arr(1) := 100
OPERATIONS
   set ( ix , tt ) =
   IF
       ix >= 0 & ix <=99 & tt >= 0  & tt <= 1000
   THEN

@@

\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{llvm}

@@

@@

\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{pascal}
MACHINE
   Bubble
VARIABLES
   vec,sort
INVARIANT
   vec : 0..99 --> 0..99 & sort : 0..1 &
   ( sort = 1 =>  !ii.(ii : 0..98 => vec(ii)<=vec(ii+1)))
INITIALISATION
   vec :: 0..99 --> 0..99 || sort := 0
OPERATIONS
   op_sort =
   ANY sorted_vector WHERE
    sorted_vector : 0..99 --> 0..99 &
    !ii.(ii : 0..98 => sorted_vector(ii)<=sorted_vector(ii+1))
   THEN
    vec := sorted_vector|| sort := 1
   END

@@
